---
title: "The 'Mapping Warbler Song' Project"
author: "Mike Allen"
date: "April 2023"
output: html_document
---
# Load libraries and read in your data file
This code file is organized into discrete "chunks" each of which can be run by clicking its green arrow. Make sure the "prawar_class_data_2023.csv" file is in your "my_data" folder. The first chunk of code reads in the compiled class song data, bins them into 2x2 latitude and longitude grid cells, and converts the units from milimeters to kHz or seconds as needed. Be sure to run this chunk of code first or the others will not run.
```{r}
# load the function libraries called "dplyr", "ggplot2", and "sf"
# note: this also installs them if they are not yet installed
if (require("dplyr") == F) {
  install.packages("dplyr")
}
if (require("ggplot2") == F) {
  install.packages("ggplot2")
}
if (require("sf") == F) {
  install.packages("sf")
}
if (require("viridis") == F) {
  install.packages("viridis")
}

library(dplyr)
library(ggplot2)
library(sf)
library(viridis)

# read in the compiled class data that you put into your "my_data" folder
  # also add map 4x4 lat/lon "grid cell" information to the data
my_data_raw <- read.csv("prawar_data/2024_class_data/prawar_class_data_2023_2024.csv") %>%
  mutate(lat = trunc(Latitude), lon = trunc(Longitude)) %>%
  mutate(grid_lat = case_when(lat %in% 26:29 ~ 28,
                              lat %in% 30:33 ~ 32,
                              lat %in% 34:37 ~ 36,
                              lat %in% 38:41 ~ 40,
                              lat %in% 42:45 ~ 44,
                              lat %in% 46:49 ~ 48),
         grid_lon = case_when(lon %in% -62:-65 ~ -64,
                              lon %in% -66:-69 ~ -68,
                              lon %in% -70:-73 ~ -72,
                              lon %in% -74:-77 ~ -76,
                              lon %in% -78:-81 ~ -80,
                              lon %in% -82:-85 ~ -84,
                              lon %in% -86:-89 ~ -88,
                              lon %in% -90:-93 ~ -92,
                              lon %in% -94:-97 ~ -96),
         grid = paste0(grid_lat, grid_lon)
                              ) %>%
  select(-lat,-lon) %>%
  select(cat_num, grid_lat, grid_lon, grid, Latitude, Longitude, quality, link, 
         num_syllables, freq_range_mm, middle_syl_height_mm, duration_mm, two_khz_mm, two_sec_mm, 
         skip_notes, group_letter, group_members) %>%
  filter(!is.na(num_syllables),
         num_syllables != "",
         num_syllables != "n/a",
         num_syllables != "-",
         num_syllables != "skip",
         num_syllables != "Skip",
         duration_mm != "") %>%
  mutate(num_syllables = as.numeric(num_syllables),
         freq_range_mm = as.numeric(freq_range_mm),
         middle_syl_height_mm = as.numeric(middle_syl_height_mm),
         duration_mm = as.numeric(duration_mm),
         two_khz_mm = as.numeric(two_khz_mm),
         two_sec_mm = as.numeric(two_sec_mm),
         freq_range_khz = 2*freq_range_mm/two_khz_mm,
         middle_syl_height_khz = 2*middle_syl_height_mm/two_khz_mm,
         duration_s = 2*duration_mm/two_sec_mm,
         trill_rate = num_syllables/duration_s,
         grp = as.numeric(as.factor(group_members)))

# this step filters out highly unusual values that were likely data collection or entry errors
  # we also average values for songs that were measured more than once
my_data_tmp <- my_data_raw %>%
  filter(duration_s < 5,
         freq_range_khz < 10,
         trill_rate < 20) %>%
  mutate(dup = 1*duplicated(cat_num)) %>%
  group_by(cat_num, grid_lat, grid_lon, grid, Latitude, Longitude, link) %>%
  summarize(n_meas = length(num_syllables),
            num_syllables_cv = sd(num_syllables)/mean(num_syllables),
         freq_range_khz_cv = sd(freq_range_khz)/mean(freq_range_khz),
         middle_syl_height_khz_cv = sd(middle_syl_height_khz, na.rm = T)/mean(middle_syl_height_khz, na.rm = T),
         duration_s_cv = sd(duration_s)/mean(duration_s),
         trill_rate_cv = sd(trill_rate)/mean(trill_rate),
         num_syllables = mean(num_syllables),
         freq_range_khz = mean(freq_range_khz),
         middle_syl_height_khz = mean(middle_syl_height_khz, na.rm = T),
         duration_s = mean(duration_s),
         trill_rate = mean(trill_rate),
         .groups = "drop")

# this puts the highly unusual values into a separate database to check on
data_to_check <- my_data_raw %>%
  filter(duration_s >= 5 |
         freq_range_khz >= 10 |
         trill_rate >= 20)

# final data file
my_data <- my_data_tmp %>%
  mutate(cv_flag = case_when(num_syllables_cv > 0.33 |
           freq_range_khz_cv > 0.33 |
           middle_syl_height_khz_cv > 0.33 |
           duration_s_cv > 0.33 |
           trill_rate_cv > 0.33 ~ 1,
           TRUE ~ 0)) %>%
  select(cat_num:link, num_syllables:trill_rate, n_meas, cv_flag)

# remove raw data
rm(my_data_tmp)

# load the map of North America
state_bounds <- readRDS("map_data/prawar_map.rds")
```
# View and explore the compiled class data
Run these lines of code one by one. Do this by putting your cursor on each line of code and hitting control (or command on Mac) + enter. Look at the output from each.
QUESTION 4 in you questions document: How many rows and columns does it have?
QUESTION 5: What is the average number of syllables? The min? The max?
```{r}
# view a snapshot of the data
str(my_data)

# view the whole data base
View(my_data)

# view the mean and quantiles of the syllable number data
summary(my_data$num_syllables)

# make a histogram of the syllable numbers and other data - a lot of variation!
# check against the values we excluded above - do you see how they would be extreme outliers?
hist(my_data$num_syllables)
hist(my_data$duration_s)
hist(my_data$freq_range_khz)
hist(my_data$trill_rate) 
```
# Plot song duration vs. number of syllables?
QUESTION 6: Do you think there is a relationship between song duration and number of syllables? Explain. If no, or if it is not a very strong relationship, what does this imply?
```{r}

# song duration vs. number of syllables
my_data %>%
ggplot(aes(x = num_syllables, y = duration_s)) +
  geom_point() +
  geom_smooth(method = "lm", color = "salmon") +
  theme_bw()

```

# How about trill rate vs. number of syllables? Trill rate vs. Latitude?
Run these three plots individually and examine the output.
Notice that I removed the extremely fast trill rate (60/s) as a probably error.
QUESTION 7: Do you think there is a relationship between trill rate and number of syllables? Explain.
QUESTION 8: Do you think there is a relationship between trill rate and Latitude? Explain.
```{r}

my_data %>%
  # filter(cv_flag == 0,
  #        n_meas > 1) %>%
  # filter(duration_s < 3) %>%
  # filter(num_syllables > 4) %>%
  ggplot(aes(y = trill_rate, x = num_syllables, color = Latitude)) +
  scale_color_viridis() +
  geom_point(size = 3, alpha = .5) +
  geom_smooth(method = "lm", color = "salmon") +
  theme_bw()

my_data %>%
  # filter(cv_flag == 0,
  #        n_meas > 1) %>%
ggplot(aes(x = Latitude, y = trill_rate)) +
  scale_color_viridis() +
  geom_point(size = 3, alpha = .5) +
  theme_bw()

my_data %>%
  # filter(cv_flag == 0,
  #        n_meas > 1) %>%
  ggplot(aes(x = Latitude, y = num_syllables)) +
  scale_color_viridis() +
  geom_point(size = 3, alpha = .5) +
  theme_bw()

my_data %>%
  # filter(cv_flag == 0,
  #        n_meas > 1) %>%
ggplot(aes(x = Latitude, y = duration_s)) +
  scale_color_viridis() +
  geom_point(size = 3, alpha = .5) +
  theme_bw()

```
# Plot number of syllables vs. frequency range
QUESTION 9: Do you think there is a relationship between number of syllables and frequency range?
QUESTION 10: Do you think there is a relationship between Latitude and frequency range?
```{r}

# number of syllables vs. frequency range
my_data %>%
ggplot(aes(x = num_syllables, y = freq_range_khz)) +
  geom_point() +
  geom_smooth(method = "lm", color = "salmon") +
  theme_bw()

my_data %>%
ggplot(aes(x = Latitude, y = freq_range_khz)) +
  geom_point() +
  geom_smooth(method = "lm", color = "salmon") +
  theme_bw()

```

# Let's do some QC  plotting 
to see if might have been systematic measuring errors by group
Note: these data are from before we filtered out "unreasonable" values
```{r}

my_data_raw %>%
  ggplot(aes(x = as.factor(grp), y = duration_s)) +
  geom_boxplot()

my_data_raw %>%
  ggplot(aes(x = as.factor(grp), y = freq_range_khz)) +
  geom_boxplot()

my_data_raw %>%
  ggplot(aes(x = as.factor(grp), y = num_syllables)) +
  geom_boxplot()

my_data_raw %>%
  # filter(grp != 13) %>%
  ggplot(aes(x = as.factor(grp), y = trill_rate)) +
  geom_boxplot()

```

# Summarize the data by grid cell
```{r}
# summaize data by grid cell
sum_data <- my_data %>%
  group_by(grid, grid_lat, grid_lon) %>%
  summarize(num_syllables = mean(num_syllables),
            freq_range_khz = mean(freq_range_khz),
            middle_syl_height_khz = mean(middle_syl_height_khz),
            duration_s = mean(duration_s),
            trill_rate = mean(trill_rate),
            n = length(grid_lat),
            .groups = "drop")

View(sum_data)
```
# Look at the averaged grid cell data in terms of song trait pairs & lattitude
This can help us see geographic clusters. QUESTION 11: Do you see any?
```{r}
sum_data %>%
  ggplot() +
  geom_point(aes(x = num_syllables, y = trill_rate, 
                 color = grid_lat), size = 4) +
  scale_color_viridis() +
  theme_bw()

```

# Make map of the mean syllables per song
Make and save Figure 1.
QUESTION 12: Do you see any patterns in mean number of syllables?
```{r}
sum_data %>%
   filter(n >= 3) %>%
    ggplot() +
  geom_sf(data = state_bounds, aes(), color = "gray",
          linewidth = .25) +
  geom_tile(aes(x = grid_lon, y = grid_lat, fill = num_syllables), 
                 size = 1) +
  scale_fill_viridis(option = "inferno") +
  theme_bw() + 
  theme(text = element_text(size = 14)) +
  labs(x = "", y = "",
       fill = "Mean\nnumber",
       title = "Number of syllables") +
  geom_point(aes(x = Longitude, y = Latitude), data = my_data, shape = "+", color = "gray")

ggsave("my_output/Fig1_mean_num_syllables.png", 
       height = 5, width = 5, dpi = 400)
```
# Make map of the mean frequency range per song
Make and save Figure 2.
QUESTION 13: Do you see any patterns in mean frequency range?
```{r}
sum_data %>%
  filter(n >= 3) %>%
    ggplot() +
  geom_sf(data = state_bounds, aes(), color = "gray",
          size = .25) +
  geom_tile(aes(x = grid_lon, y = grid_lat, fill = freq_range_khz), 
                 size = 1) +
  scale_fill_viridis(option = "inferno") +
  theme_bw() + 
  theme(text = element_text(size = 14)) +
  labs(x = "", y = "", fill = "Mean\nrange\n(kHz)",
       title = "Frequency range") +
  geom_point(aes(x = Longitude, y = Latitude), data = my_data, shape = "+", color = "gray")

ggsave("my_output/Fig2_mean_freq_range_khz.png", 
       height = 5, width = 5, dpi = 400)
```
# Make map of the mean song duration
Make and save Figure 3.
QUESTION 14: Do you see any patterns in mean song duration?
```{r}
sum_data %>%
  filter(n >= 3) %>%
    ggplot() +
  geom_sf(data = state_bounds, aes(), color = "gray",
          size = .25) +
  geom_tile(aes(x = grid_lon, y = grid_lat, fill = duration_s), 
                 size = 1) +
  scale_fill_viridis(option = "inferno") +
  theme_bw() + 
  theme(text = element_text(size = 14)) +
  labs(x = "", y = "", fill = "Mean\nduration\n(s)",
       title = "Song duration")

ggsave("prawar_data/2024_class_output/Fig3_mean_song_duration_s.png", 
       height = 5, width = 5, dpi = 400)
```
# Challenge: make a map of the mean trill rate (syllables / second) and middle syllable height (kHz)
Modify the code below (copied from the previous chunk) to make the 4th and 5th "challenge" maps. Save these as Fig. 4 and 5. 
QUESTION 15: Do you see any patterns in mean trill rate or middle syllable height?
```{r}

sum_data %>%
  filter(n >= 3) %>%
    ggplot() +
  geom_sf(data = state_bounds, aes(), color = "gray",
          size = .25) +
  geom_tile(aes(x = grid_lon, y = grid_lat, fill = duration_s), 
                 size = 1) +
  scale_fill_viridis(option = "inferno") +
  theme_bw() + 
  theme(text = element_text(size = 14)) +
  labs(x = "", y = "", fill = "Mean\nduration\n(s)",
       title = "Song duration")

ggsave("prawar_data/2024_class_output/Fig4_mean_trill_rate.png", 
       height = 5, width = 5, dpi = 400)

ggsave("prawar_data/2024_class_output/Fig5_mean_middle_syl_height_kHz.png", 
       height = 5, width = 5, dpi = 400)
```

Go back to the README.md file on the "Mapping Warblers" GitHub page to finish up Phase 3. https://github.com/mikeallen-eco/mapping_warblers/blob/main/README.md