---
title: "The 'Mapping Warbler Song' Project - part 2"
author: "Mike Allen"
date: "April 2023"
output: html_document
---
# Load libraries and read in your data file
This code file is organized into discrete "chunks" each of which can be run by clicking its green arrow. Make sure there is only one csv file (the csv you downloaded from Macaulay Library) in the "prawar_data" folder. You'll need to delete the Prairie Warbler example file that is currently there. This code reads in the data, keeping only songs with a known location (latitude/longitude) and with a quality score of 4 or greater.
```{r}
# load the function libraries called "dplyr", "ggplot2", and "sf"
# note: this also installs them if they are not yet installed
if (require("dplyr") == F) {
  install.packages("dplyr")
}
if (require("ggplot2") == F) {
  install.packages("ggplot2")
}
if (require("sf") == F) {
  install.packages("sf")
}

library(dplyr)
library(ggplot2)
library(sf)

# read in the compiled class data that you put into your "my_output" folder
# this selects just the data from the main part of their range
      # with a quality rating of 3 or above
      # and that occur in the months of June or July

# load and filter the raw data file
my_data <- read.csv("my_output/prawar_class_data.csv")

# load the map of North America
state_bounds <- readRDS("map_data/prawar_map.rds")
```
# View and explore the compiled class data
QUESTION 1: How many rows and columns does it have?
```{r}
View(my_data)
hist(my_data$)
```
# Make a rough map of the raw data
QUESTION 2: Where are most of the song locations concentrated geographically?
```{r}
my_data %>%
    ggplot() +
  geom_sf(data = state_bounds, aes(), color = "gray",
          size = .25) +
  geom_point(aes(x = Longitude, y = Latitude), 
                 size = 1,
                 color = "firebrick", 
            alpha = .5) +
  theme_bw() + 
  theme(text = element_text(size = 14)) +
  labs(x = "", y = "")
```
# Plotting the results
Use this code to plot your results. It will save a png file to the "my_output" folder. Before you run it, change the 2nd and 3rd lines of the code to reflect your correct units and species. After you run this chunk of code, go ahead and complete the remaining sections of code in this file (described in README steps 15-17).
# First input the correct measurement units and species
Change the quoted text to reflect your own study.
```{r}
your_units <- "Number of syllables in trill"
your_species <- "Prairie Warbler"
```
# Make the plots
Run this chunk (click the green arrow) to make and save the plots. They should appear in your "my_output".
```{r}
# make sure you have the required packages just in case
if(require(ggplot2)==F){install.packages("ggplot2")
  library(ggplot2)}
if(require(dplyr)==F){install.packages("dplyr")
  library(dplyr)}

# check if output files exist
lat_exists <- 
  file.exists("my_output/my_complete_data_file_latitude.csv")
lon_exists <- 
  file.exists("my_output/my_complete_data_file_longitude.csv")

# read in the completed latitude data file
if(lat_exists){
data_complete_lat <- read.csv(
  "my_output/my_complete_data_file_latitude.csv"
  )
}

# read in the completed longitude data file
if(lon_exists){
data_complete_lon <- read.csv(
  "my_output/my_complete_data_file_longitude.csv"
  )
}

# combine the two data files (if they exist)
if(lat_exists & lon_exists){
  data_complete <- 
    data_complete_lat %>%
    bind_rows(data_complete_lon)
}

if(lat_exists==T & lon_exists==F){
  data_complete <- data_complete_lat}

if(lat_exists==F & lon_exists==T){
  data_complete <- data_complete_lon}

# plot the measurement by latitude
data_complete %>%
  ggplot(aes(x = Latitude, y = measurement)) +
  geom_point(size = 4) +
  geom_smooth(method = "loess", se = F, color = "firebrick") +
  geom_smooth(method = "lm", se = T, color = "darkgray", lty = 2) +
  theme_bw() + 
  theme(text = element_text(size = 14)) +
  labs(y = your_units, title = your_species)

# save the latitude plot
ggsave("my_output/my_plot_latitude.png", 
       height = 4, width = 6, dpi = 100)

# plot the measurement by longitude
data_complete %>%
  ggplot(aes(x = Longitude, y = measurement)) +
  geom_point(size = 4) +
  geom_smooth(method = "loess", se = F, color = "firebrick") +
  geom_smooth(method = "lm", se = T, color = "darkgray", lty = 2) +
  theme_bw() + 
  theme(text = element_text(size = 14)) +
  labs(y = your_units, title = your_species)

# save the longitude plot
ggsave("my_output/my_plot_longitude.png", 
       height = 4, width = 6, dpi = 100)

```
# Map the results
```{r}
# make sure you have the required packages installed
if(require(rnaturalearth)==F){install.packages("rnaturalearth")}
if(require(sf)==F){install.packages("sf")}
if(require(viridis)==F){install.packages("viridis")}
if(require(ggplot2)==F){install.packages("ggplot2")}
if(require(dplyr)==F){install.packages("dplyr")}
library(rnaturalearth)
library(sf)
library(viridis)
library(ggplot2)
library(dplyr)

# make the data into a "spatial" format
data_spatial <- data_complete %>%
  mutate(lat = Latitude, lon = Longitude) %>%
  sf::st_as_sf(coords = c("Longitude", "Latitude"), 
               crs = 4326) 

# load the continental boundaries
state_bounds <- ne_coastline(scale = 10, returnclass = "sf") %>%
  st_crop(st_buffer(data_spatial, dist = 2000000))

# make the map
data_spatial %>%
  ggplot() +
  geom_point(aes(x = lon, y = lat, 
                 size = measurement,
                 color = measurement), 
            alpha = .75) +
  viridis::scale_color_viridis(option = "magma") +
  geom_sf(data = state_bounds, aes(), color = "gray",
          size = .25) + 
  theme_bw() + 
  theme(text = element_text(size = 14)) +
  labs(size = your_units, title = your_species,
       x = "", y = "", color = your_units)

# save the map
ggsave("my_output/my_map.png", 
       height = 6, width = 7, dpi = 100)

```
# Describe your methods
Replace the DESCRIBE YOUR METHODS HERE text with your own. Don't delete any of the quotation marks or other text. Be sure to clearly describe what you measured and how. Write as much as you need to, but only replace the text within the asterisks(*). After you are done writing, run the code to write a text file using the green arrow.
```{r}
cat(file = "my_output/my_methods.txt", 
    "
METHODS

*DESCRIBE YOUR METHODS HERE*


    "
)

```
# Describe your results
Replace the DESCRIBE YOUR RESULTS HERE text with your own. Don't delete any of the quotation marks or other text. Write as much as you feel you need to, but only replace the text within the asterisks(*). After you are done writing, run the code to write a text file using the green arrow.
```{r}
cat(file = "my_output/my_results.txt", 
    "
RESULTS

*DESCRIBE YOUR RESULTS HERE*


    "
)
```
