---
title: "The 'Mapping Warblers' Project"
author: "Mike Allen"
date: "1/31/2022"
output: html_document
---
# Load libraries and read in your data file
This code file is organized into discrete "chunks" each of which can be run by clicking its green arrow. Make sure there is only one file (the csv you downloaded from Macaulay Library) in the "my_raw_data" folder. You'll need to delete the Prairie Warbler example file that is currently there. This code reads in the data, keeping only songs with a known location (latitude/longitude) and with a quality score of 4 or greater.
```{r}
# load the function libraries called "dplyr" and "ggplot2"
# note: this also installs them if they are not yet installed
if (require("dplyr") == F) {
  install.packages("dplyr")
}
if (require("ggplot2") == F) {
  install.packages("ggplot2")
}
if (require("sf") == F) {
  install.packages("sf")
}

library(dplyr)
library(ggplot2)
library(sf)

# read in the data from your "prawar_data" folder
# note: it also prints a warning if you have >1 file in your folder
if (length(list.files("prawar_data")) != 1) {
  warning(
    "Only 1 file can be in the prawar_data folder. Please delete any except the one you want to analyze."
  )
} else{
  my_data <- read.csv(paste0("prawar_data/",
                             list.files("prawar_data"))) %>%
    filter(Average.Community.Rating >= 3,
           is.na(Latitude) != TRUE,
           Latitude > 30,
           Longitude > -98,
           Month %in% c(6,7)) %>%
    rename(cat_num = 1)
}

# load the map of North America
state_bounds <- readRDS("map_data/prawar_map.rds")
```
# View and explore the filtered raw data
QUESTION 1: How many rows and columns does it have?
```{r}
View(my_data)
```
# Make a rough map of the raw data
QUESTION 2: Where are most of the song locations concentrated geographically?
```{r}
my_data %>%
    ggplot() +
  geom_sf(data = state_bounds, aes(), color = "gray",
          size = .25) +
  geom_point(aes(x = Longitude, y = Latitude), 
                 size = 1,
                 color = "firebrick", 
            alpha = .5) +
  theme_bw() + 
  theme(text = element_text(size = 14)) +
  labs(x = "", y = "")
```
# calculate number of latitude bands your data spans
QUESTION 3: Which latitude band has the most song recordings?
QUESTION 4: Which state has the most song recordings?
```{r}
# print how many latitude and longitude bands the dataset covers
length(unique(floor(my_data$Latitude)))
length(unique(floor(my_data$Longitude)))

# print how many songs are in each latitude or longitude band
table(floor(my_data$Latitude))
table(floor(my_data$Longitude))

#print how many song recordings are available for each state
table(my_data$State)
```
# Create final the data file
CHANGE THE GROUP NAME AND SONG NUMBERS TO MATCH YOUR GROUP. Then, hit the green arrow to run this code that creates a final blank data file for you. You will first need to input your group name and the range of song numbers you were assigned at the top of the chunk. Once you do this, move the file from the "my_output" folder to the "my_data" folder.
```{r}
# WHAT IS YOUR GROUP NAME? (keep it short, no spaces allowed)
# AND WHAT SONG NUMBERS WERE YOU ASSIGNED TO MEASURE?
  # change the xx-xx to those numbers (e.g., 1:50)
my_group_name <- "group_name"
my_song_numbers <- 1:50

# create a randomized data set tt
my_data_final <- my_data %>%
  mutate(song_num = 1:nrow(my_data)) %>%
  filter(song_num %in% my_song_numbers) %>%
  select(cat_num, Latitude, Longitude, 
         quality = Average.Community.Rating) %>%
  mutate(link = paste0("https://macaulaylibrary.org/asset/", 
                       cat_num),
         num_sylables = "", 
         trill_length = "",
         num_harmonics = "", 
         skip_notes = ""
)

# write the final "blank" data file to csv files

write.csv(my_data_final, 
          paste0("my_output/", my_group_name, "_data_file.csv"),
          row.names = F)

```
# Making your measurements
At this point, minimize RStudio, go back to the README file, and complete steps 9-14. In those steps, you'll use spreadsheet software to open the "my_data_file.csv" file you just created. Then, you'll go to each link, view the sonogram, and make the measurements you came up with using a ruler or visual observation. (See README file for more info.) You'll add those measurements to your data file and then come back here to graph the results.

# Graph the results
Use this code to plot your results. It will save a png file to the "my_output" folder. Before you run it, change the 2nd and 3rd lines of the code to reflect your correct units and species. After you run this chunk of code, go ahead and complete the remaining sections of code in this file (described in README steps 15-17).
# First input the correct measurement units and species
Change the quoted text to reflect your own study.
```{r}
your_units <- "Number of syllables in trill"
your_species <- "Prairie Warbler"
```
# Make the plots
Run this chunk (click the green arrow) to make and save the plots. They should appear in your "my_output".
```{r}
# make sure you have the required packages just in case
if(require(ggplot2)==F){install.packages("ggplot2")
  library(ggplot2)}
if(require(dplyr)==F){install.packages("dplyr")
  library(dplyr)}

# check if output files exist
lat_exists <- 
  file.exists("my_output/my_complete_data_file_latitude.csv")
lon_exists <- 
  file.exists("my_output/my_complete_data_file_longitude.csv")

# read in the completed latitude data file
if(lat_exists){
data_complete_lat <- read.csv(
  "my_output/my_complete_data_file_latitude.csv"
  )
}

# read in the completed longitude data file
if(lon_exists){
data_complete_lon <- read.csv(
  "my_output/my_complete_data_file_longitude.csv"
  )
}

# combine the two data files (if they exist)
if(lat_exists & lon_exists){
  data_complete <- 
    data_complete_lat %>%
    bind_rows(data_complete_lon)
}

if(lat_exists==T & lon_exists==F){
  data_complete <- data_complete_lat}

if(lat_exists==F & lon_exists==T){
  data_complete <- data_complete_lon}

# plot the measurement by latitude
data_complete %>%
  ggplot(aes(x = Latitude, y = measurement)) +
  geom_point(size = 4) +
  geom_smooth(method = "loess", se = F, color = "firebrick") +
  geom_smooth(method = "lm", se = T, color = "darkgray", lty = 2) +
  theme_bw() + 
  theme(text = element_text(size = 14)) +
  labs(y = your_units, title = your_species)

# save the latitude plot
ggsave("my_output/my_plot_latitude.png", 
       height = 4, width = 6, dpi = 100)

# plot the measurement by longitude
data_complete %>%
  ggplot(aes(x = Longitude, y = measurement)) +
  geom_point(size = 4) +
  geom_smooth(method = "loess", se = F, color = "firebrick") +
  geom_smooth(method = "lm", se = T, color = "darkgray", lty = 2) +
  theme_bw() + 
  theme(text = element_text(size = 14)) +
  labs(y = your_units, title = your_species)

# save the longitude plot
ggsave("my_output/my_plot_longitude.png", 
       height = 4, width = 6, dpi = 100)

```
# Map the results
```{r}
# make sure you have the required packages installed
if(require(rnaturalearth)==F){install.packages("rnaturalearth")}
if(require(sf)==F){install.packages("sf")}
if(require(viridis)==F){install.packages("viridis")}
if(require(ggplot2)==F){install.packages("ggplot2")}
if(require(dplyr)==F){install.packages("dplyr")}
library(rnaturalearth)
library(sf)
library(viridis)
library(ggplot2)
library(dplyr)

# make the data into a "spatial" format
data_spatial <- data_complete %>%
  mutate(lat = Latitude, lon = Longitude) %>%
  sf::st_as_sf(coords = c("Longitude", "Latitude"), 
               crs = 4326) 

# load the continental boundaries
state_bounds <- ne_coastline(scale = 10, returnclass = "sf") %>%
  st_crop(st_buffer(data_spatial, dist = 2000000))

# make the map
data_spatial %>%
  ggplot() +
  geom_point(aes(x = lon, y = lat, 
                 size = measurement,
                 color = measurement), 
            alpha = .75) +
  viridis::scale_color_viridis(option = "magma") +
  geom_sf(data = state_bounds, aes(), color = "gray",
          size = .25) + 
  theme_bw() + 
  theme(text = element_text(size = 14)) +
  labs(size = your_units, title = your_species,
       x = "", y = "", color = your_units)

# save the map
ggsave("my_output/my_map.png", 
       height = 6, width = 7, dpi = 100)

```
# Describe your methods
Replace the DESCRIBE YOUR METHODS HERE text with your own. Don't delete any of the quotation marks or other text. Be sure to clearly describe what you measured and how. Write as much as you need to, but only replace the text within the asterisks(*). After you are done writing, run the code to write a text file using the green arrow.
```{r}
cat(file = "my_output/my_methods.txt", 
    "
METHODS

*DESCRIBE YOUR METHODS HERE*


    "
)

```
# Describe your results
Replace the DESCRIBE YOUR RESULTS HERE text with your own. Don't delete any of the quotation marks or other text. Write as much as you feel you need to, but only replace the text within the asterisks(*). After you are done writing, run the code to write a text file using the green arrow.
```{r}
cat(file = "my_output/my_results.txt", 
    "
RESULTS

*DESCRIBE YOUR RESULTS HERE*


    "
)
```
