---
title: "The 'Mapping Warbler Song' Project - part 2"
author: "Mike Allen"
date: "April 2023"
output: html_document
---
# Load libraries and read in your data file
This code file is organized into discrete "chunks" each of which can be run by clicking its green arrow. Make sure the "prawar_class_data_2023.csv" file is in your "my_data" folder. The first chunk of code reads in the compiled class data, keeping only songs with a known location (latitude/longitude) and with a quality score of 4 or greater.
```{r}
# load the function libraries called "dplyr", "ggplot2", and "sf"
# note: this also installs them if they are not yet installed
if (require("dplyr") == F) {
  install.packages("dplyr")
}
if (require("ggplot2") == F) {
  install.packages("ggplot2")
}
if (require("sf") == F) {
  install.packages("sf")
}
if (require("viridis") == F) {
  install.packages("viridis")
}

library(dplyr)
library(ggplot2)
library(sf)
library(viridis)

# create a "my_data" folder if none exists
if(!dir.exists("my_data")){dir.create("my_data")}

# read in the compiled class data that you put into your "my_data" folder
  # also add map 4x4 lat/lon "grid cell" information to the data
my_data <- read.csv("my_data/prawar_class_data_2023.csv") %>%
  mutate(lat = trunc(Latitude), lon = trunc(Longitude)) %>%
  mutate(grid_lat = case_when(lat %in% 30:33 ~ 32,
                              lat %in% 34:37 ~ 36,
                              lat %in% 38:41 ~ 40,
                              lat %in% 42:45 ~ 44,
                              lat %in% 46:49 ~ 48),
         grid_lon = case_when(lon %in% -66:-69 ~ -68,
                              lon %in% -70:-73 ~ -72,
                              lon %in% -74:-77 ~ -76,
                              lon %in% -78:-81 ~ -80,
                              lon %in% -82:-85 ~ -84,
                              lon %in% -86:-89 ~ -88,
                              lon %in% -90:-93 ~ -92,
                              lon %in% -94:-97 ~ -96),
         grid = paste0(grid_lat, grid_lon)
                              ) %>%
  select(-lat,-lon) %>%
  select(cat_num, grid_lat, grid_lon, grid, Latitude, Longitude, quality, link, 
         num_syllables, freq_range_mm, duration_mm, two_khz_mm, two_sec_mm, 
         skip_notes, group_letter, group_members) %>%
  filter(!is.na(num_syllables),
         num_syllables != "",
         num_syllables != "n/a",
         duration_mm != "") %>%
  mutate(num_syllables = as.numeric(num_syllables),
         freq_range_mm = as.numeric(freq_range_mm),
         duration_mm = as.numeric(duration_mm),
         two_khz_mm = as.numeric(two_khz_mm),
         two_sec_mm = as.numeric(two_sec_mm),
         freq_range_khz = 2*freq_range_mm/two_khz_mm,
         duration_s = 2*duration_mm/two_sec_mm,
         trill_rate = num_syllables/duration_s,
         grp = as.numeric(as.factor(group_members)))

# load the map of North America
state_bounds <- readRDS("map_data/prawar_map.rds")
```
# View and explore the compiled class data
QUESTION 5: How many rows and columns does it have?
QUESTION 6: What is the average number of syllables? The min? The max?
```{r}
# view a snapshot of the data
str(my_data)

# view the whole data base
View(my_data)

# view the mean and quantiles of the syllable number data
summary(my_data$num_syllables)

# make a histogram of the syllable number data - a lot of variation!
hist(my_data$num_syllables)
hist(my_data$duration_s)
hist(my_data$freq_range_khz)
hist(my_data$trill_rate) # do you see an outlier here?
```
# Do you think there is a relationship between song duration and number of syllables?
What do you think?
```{r}

# song duration vs. number of syllables
my_data %>%
ggplot(aes(x = num_syllables, y = duration_s)) +
  geom_point() +
  geom_smooth(method = "lm", color = "salmon") +
  theme_bw()

```

# How about trill rate vs. number of syllables?
Notice that I removed the extremely fast trill rate (60/s) as a probably error.
```{r}

my_data %>%
  filter(trill_rate < 60) %>%
ggplot(aes(y = trill_rate, x = num_syllables)) +
  geom_point() +
  geom_smooth(method = "lm", color = "salmon") +
  theme_bw()

```
# Do you think there is a relationship between number of syllables and frequency range?
What do you think?
```{r}

# number of syllables vs. frequency range
my_data %>%
ggplot(aes(x = num_syllables, y = freq_range_khz)) +
  geom_point() +
  geom_smooth(method = "lm", color = "salmon") +
  theme_bw()

```

# Let's do some QC  plotting 
to see if might have been systematic measuring errors by group
```{r}

my_data %>%
  ggplot(aes(x = as.factor(grp), y = duration_s)) +
  geom_boxplot()

my_data %>%
  ggplot(aes(x = as.factor(grp), y = freq_range_khz)) +
  geom_boxplot()

my_data %>%
  ggplot(aes(x = as.factor(grp), y = num_syllables)) +
  geom_boxplot()

my_data %>%
  ggplot(aes(x = as.factor(grp), y = trill_rate)) +
  geom_boxplot()

```

# Summarize the data by grid cell
Notice that I filter out a data point with an extremely high trill rate
```{r}
# summaize data by grid cell
sum_data <- my_data %>%
  filter(trill_rate < 60) %>%
  group_by(grid, grid_lat, grid_lon) %>%
  summarize(num_syllables = mean(num_syllables),
            freq_range_khz = mean(freq_range_khz),
            duration_s = mean(duration_s),
            trill_rate = mean(trill_rate),
            n = length(grid_lat),
            .groups = "drop")

View(sum_data)
```
# Look at the averaged grid cell data in terms of song trait pairs & lattitude
This can help us see geographic clusters. Do you see any?
```{r}
sum_data %>%
  ggplot() +
  geom_point(aes(x = num_syllables, y = trill_rate, 
                 color = grid_lat), size = 4) +
  scale_color_viridis() +
  theme_bw()

```

# Make map of the mean syllables per song
QUESTION 7: Do you see any patterns in mean number of syllables?
```{r}
sum_data %>%
  # filter(n >= 3) %>%
    ggplot() +
  geom_sf(data = state_bounds, aes(), color = "gray",
          size = .25) +
  geom_tile(aes(x = grid_lon, y = grid_lat, fill = num_syllables), 
                 size = 1) +
  scale_fill_viridis(option = "inferno") +
  theme_bw() + 
  theme(text = element_text(size = 14)) +
  labs(x = "", y = "",
       fill = "Mean\nnumber",
       title = "Number of syllables")

ggsave("my_output/Fig1_mean_num_syllables.png", 
       height = 5, width = 5, dpi = 400)
```
# Make map of the mean frequency range per song
QUESTION 8: Do you see any patterns in mean frequency range?
```{r}
sum_data %>%
  # filter(n >= 3) %>%
    ggplot() +
  geom_sf(data = state_bounds, aes(), color = "gray",
          size = .25) +
  geom_tile(aes(x = grid_lon, y = grid_lat, fill = freq_range_khz), 
                 size = 1) +
  scale_fill_viridis(option = "inferno") +
  theme_bw() + 
  theme(text = element_text(size = 14)) +
  labs(x = "", y = "", fill = "Mean\nrange\n(kHz)",
       title = "Frequency range")

ggsave("my_output/Fig2_mean_freq_range_khz.png", 
       height = 5, width = 5, dpi = 400)
```
# Make map of the mean song duration
QUESTION 10: Do you see any patterns in mean song duration?
```{r}
sum_data %>%
  # filter(n >= 3) %>%
    ggplot() +
  geom_sf(data = state_bounds, aes(), color = "gray",
          size = .25) +
  geom_tile(aes(x = grid_lon, y = grid_lat, fill = duration_s), 
                 size = 1) +
  scale_fill_viridis(option = "inferno") +
  theme_bw() + 
  theme(text = element_text(size = 14)) +
  labs(x = "", y = "", fill = "Mean\nduration\n(s)",
       title = "Song duration")

ggsave("my_output/Fig3_mean_song_duration_s.png", 
       height = 5, width = 5, dpi = 400)
```
# Challenge: make a map of the mean trill rate (syllables / second)
Copy / paste and modify the code in the previous chunk to make the 4th "challenge" map. Save this as Fig. 4. Do you see any patterns in mean trill rate?
```{r}


```

Go back to the "Mapping Warblers" GitHub page to finish up Phase 3. 